---
title: "Using and validating LAMBclassifier"
author: "Debjeet Ray"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using and validating LAMBclassifier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
# Chunks that need no external data are individually marked eval = TRUE.
```

## Scope

LAMBclassifier assigns bulk expression profiles to six expression-defined
classes using a locked, deterministic, single-sample rank-scoring rule. It is a
research tool. A call is not a pathology diagnosis, a receptor assay, or a
substitute for PAM50.

The training set is an in-house PDX cohort. It is the sole training set, and it
cannot be published yet. TCGA was not training data; it contributed only to gene
availability in the fixed scoring universe, so TCGA results are characterization
rather than fully independent validation of the complete pipeline.

## Classification

The input is a numeric matrix with HGNC gene symbols in rows and samples in
columns. Use comparable sample-level normalization and a log2-like expression
scale; do not normalize a new validation cohort together with any development
cohort.

```{r classify}
library(LAMBclassifier)

expr <- read.delim(
  "expression_matrix.txt", row.names = 1, check.names = FALSE
)
expr <- prepare_lamb_data(expr, log2_transform = FALSE)

# Set run_pam50 = FALSE when PAM50 is unavailable or not part of the protocol.
results <- LAMBclassifier(expr, run_pam50 = TRUE)
summary_lamb_results(results)

lamb_specification()
export_results(results, "results/classification_results.csv")
```

`export_results()` writes the result CSV plus provenance and ordered-signature
TSVs. The main output includes `Pred_Class`, `Margin`, `Flag`, all fixed
signature scores and coverages, and specification hashes. `close_call`,
`low_coverage`, and `NC` are retained in primary analyses.

Custom signatures create a different classifier and require a new development
and validation cycle. They can be supplied explicitly, for example:

```{r custom}
custom_results <- LAMBclassifier(
  expr,
  chondroid_genes = c("SP7", "ACAN", "COL2A1"),
  squamous_genes = c("TP63", "KRT5", "KRT14"),
  basal_like_genes = c("FOXC1", "SOX10", "ELF5"),
  lhs_genes = c("ESR1", "GATA3", "PGR"),
  apocrine_genes = c("AR", "FOXA1", "FKBP5"),
  malo_her2_genes = c("ERBB2", "GRB7", "STARD3")
)
```

## A runnable example

The chunks below that read files are not evaluated, because no expression matrix
ships with the package. This one is, so you can see the shape of a result
without any data of your own.

```{r demo, eval = TRUE}
library(LAMBclassifier)

# A toy bulk-like matrix: 3,000 genes drawn from the reference universe.
set.seed(1)
genes <- unique(c(DEFAULT_ALL_INPUT_GENES, sample(lamb_universe(), 3000)))
expr <- matrix(
  stats::runif(length(genes) * 3),
  nrow = length(genes),
  dimnames = list(genes, c("tumour_A", "tumour_B", "tumour_C"))
)

# Push one sample towards the basal-like signature.
expr[DEFAULT_BASAL_LIKE_GENES, "tumour_B"] <- 5

res <- LAMBclassifier(expr, run_pam50 = FALSE)
res[, c("Sample", "Pred_Class", "Margin", "Flag", "Unscored_Signatures")]
```

Every row also carries the specification version and both hashes, so a result
can always be traced to the classifier that produced it:

```{r demo-provenance, eval = TRUE}
res[1, c("Classifier_Version", "Signature_SHA256", "Universe_SHA256")]
```

Signature coverage is reported per signature, which is what makes targeted
panels safe to use:

```{r demo-coverage, eval = TRUE}
scores <- lamb_score(expr)
attr(scores, "coverage")
```

## Single-cell and spatial data

There is no separate single-cell version. Because `singscore` ranks genes within
one sample, a cluster pseudobulk scores on exactly the same scale as a bulk
tumour — so the workflow is: aggregate to pseudobulk, then classify.

```{r singlecell}
# Seurat: sum raw counts per cluster, then CPM-normalise and log2.
counts <- Seurat::GetAssayData(seurat_obj, layer = "counts")
clusters <- seurat_obj$seurat_clusters

pseudobulk <- vapply(
  levels(clusters),
  function(cl) Matrix::rowSums(counts[, clusters == cl, drop = FALSE]),
  numeric(nrow(counts))
)

cpm <- t(t(pseudobulk) / colSums(pseudobulk)) * 1e6
log_cpm <- log2(cpm + 1)

cluster_calls <- LAMBclassifier(log_cpm, run_pam50 = FALSE)
```

Classify each cluster, then weight by cell count to describe a whole sample as a
composition rather than forcing it into one label. Averaging a heterogeneous
slide into a single pseudobulk produces a profile that belongs to no real cell
population, so per-cluster calls are the meaningful unit.

```{r composition}
cells_per_cluster <- table(clusters)
composition <- data.frame(
  Class = cluster_calls$Pred_Class,
  Cells = as.integer(cells_per_cluster[cluster_calls$Sample])
)
aggregate(Cells ~ Class, composition, sum)
```

Single-cell and spatial calls are **not externally validated**. No such cohort
with independent class labels has been evaluated, so treat these calls as
exploratory.

A panel must carry enough genes to evaluate every gate. Coverage is tracked per
signature; a signature with fewer than `min_genes_per_sig` present genes is left
unscored, and any sample whose decision path needed it returns `NC` with the
`incomplete_gates` flag. Scoring is refused outright below 100
reference-universe genes. Check `Flag`, `Unscored_Signatures` and the coverage
attribute before interpreting.

## Plots

All plotting functions take the result of `LAMBclassifier()` and write files to
`output_prefix`.

```{r plots}
# Signature heatmap, samples annotated by class
plot_comprehensive_heatmap(
  expr_mat = expr, results = res, output_prefix = "figures/heatmap"
)

# Class counts arranged on the mammary differentiation hierarchy
plot_lineage_tree(res, output_prefix = "figures/lineage_tree")

# Pairwise score spaces with the y = x decision boundary drawn in
plot_decision_boundaries(res, output_prefix = "figures/boundaries")

# Two genes against each other, coloured by call
plot_expression_scatter(expr, res, "GATA3", "AR", output_prefix = "figures/gata3_ar")

# Cross-classification against any reference column present in `res`
plot_confusion_matrix(res, truth_col = "PAM50", output_prefix = "figures/confusion")

# Convenience wrapper: the standard heatmap set in one call
generate_standard_heatmaps(
  expr_mat = expr, results = res, output_prefix = "figures/standard"
)
```

## PAM50 comparison

PAM50 and LAMB describe different taxonomies. Report their full
cross-classification and association rather than calling PAM50 ground truth for
six-class accuracy.

```{r pam50}
tab_counts <- lamb_cross_tab(results$Pred_Class, results$PAM50)
tab_by_lamb <- lamb_cross_tab(
  results$Pred_Class, results$PAM50, normalize = "prediction"
)
association <- lamb_cramers_v(results$Pred_Class, results$PAM50)

lamb_binary_metrics(
  predicted = results$Pred_Class,
  reference = results$PAM50,
  positive_predicted = "LHS",
  positive_reference = c("LumA", "LumB")
)
```

## Clinical concordance

Join annotations by stable sample identifier and evaluate each endpoint only
where that reference is known. Declare the positive prediction and clinical
state explicitly.

```{r clinical}
clinical <- read.csv("clinical_annotations.csv", stringsAsFactors = FALSE)
dat <- merge(results, clinical, by = "Sample", all.x = TRUE)

er_concordance <- lamb_binary_metrics(
  dat$Pred_Class, dat$ER_status,
  positive_predicted = "LHS", positive_reference = "Positive"
)
her2_concordance <- lamb_binary_metrics(
  dat$Pred_Class, dat$HER2_status,
  positive_predicted = "MA-Lo (HER2)", positive_reference = "Positive"
)
tn_concordance <- lamb_binary_metrics(
  dat$Pred_Class, dat$Triple_negative,
  positive_predicted = "Basal-like", positive_reference = TRUE
)
```

These mappings assess concordance. MA-Lo is ERBB2-amplicon expression, not a
DNA-amplification or clinical HER2 test. Chondroid/Squamous require pathology
labels for histologic validation, and Apocrine requires independent ER and AR
evidence for the corresponding biological claim.

## Independent-validation safeguards

Freeze the classifier, preprocessing, eligibility criteria, endpoints, class
mappings, and analysis code before accessing outcomes. Keep cohorts separate,
check patient overlap, use biological patients/models as independent units,
reserve technical replicates for repeatability, and report all evaluable calls
and 2x2 cells with confidence intervals. Never tune the classifier on a cohort
then continue to call that same cohort independent validation.

See `system.file("METHODS.md", package = "LAMBclassifier")` for the full policy.
